Analysis_v1.1_weather

### 2022-01-09 R/Analysis_v1.1_weather_2020.Rmd
Add hypothesis matrices using weather data from 3 years

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(here)
library(zoo)
```

@@

  arrange(Latitude, Year, JDAY) %>%

@@

GR_start <- phe_gr_fl %>% 
  group_by(manu_site, Year) %>%
  filter(!is.na(GR50)) %>%
  summarise(GR_start = floor(min(GR50)))
FL_end <- phe_gr_fl %>% 
  group_by(manu_site, Year) %>%
  filter(!is.na(FL50)) %>%
  summarise(FL_end = ceiling(max(FL50)))

tmp1 <- wea_ten_sites %>%
  mutate(JDAY_EXP = case_when(Year == 2018 ~ JDAY,
                              Year == 2019 ~ JDAY + 365,
                              Year == 2020 ~ JDAY + 365 + 365,
                              Year == 2021 ~ JDAY + (365*3),
                              TRUE ~ NA_real_)) %>%
  left_join(phe_gr_fl)

@@

tmp %>%
  filter(manu_site == "TX1")
tmp
```

```{r}
phe_ten_sites %>%
  left_join(wea_ten_sites)
```

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(here)
library(zoo)
library(cowplot)
```

@@

  mutate(JDAY_EXP = case_when(Year == 2018 ~ JDAY,
                              Year == 2019 ~ JDAY + 365,
                              Year == 2020 ~ JDAY + (365*2),
                              Year == 2021 ~ JDAY + (365*3),
                              TRUE ~ NA_real_)) %>%
  arrange(Latitude, Year, JDAY_EXP) %>%
  select(-JDAY) %>%

@@

tmp1 <- wea_ten_sites %>%
  left_join(phe_gr_fl)

@@

tmp %>%
  filter(manu_site == "TX1")
tmp %>%
  filter(!is.na(PLANT_ID))

```{r}
cphe_gr2fl <- tmp %>%
  filter(!is.na(PLANT_ID)) %>%
  rename(cgdd_12c_gr2fl = GDD_12Cbase, crain_gr2fl = prcp)

wea_ten_sites %>%
  filter(manu_site == "TX1" & JDAY_EXP == 715)

tmp2 <- wea_ten_sites %>%
  select(-Year)

gr_ten_sites <- phe_gr_fl %>%
  mutate(GR50 = floor(GR50)) %>%
  filter(!is.na(GR50)) %>%
  left_join(tmp2, by = c("manu_site", "GR50" = "JDAY_EXP")) %>%
  select(PLANT_ID, Year, manu_site, GR50, cgdd_12c_5d:tave_20d)

fl_ten_sites <- phe_gr_fl %>%
  filter(!is.na(FL50)) %>%
  mutate(FL50 = floor(FL50),
         GR50 = floor(GR50)) %>%
  left_join(tmp2, by = c("manu_site", "FL50" = "JDAY_EXP")) %>%
  select(PLANT_ID, Year, manu_site, GR50, FL50, prcp, DYLN, crain_2d:dyln_14d_prior) %>%
  left_join(cphe_gr2fl) %>%
  rename(crain_1d = prcp, dyln_fl50 = DYLN)
# 11 September flowering dates in 2021 at Pickle have NA's for weather values. Decided to let this be.

1354-(365*3)
tmp2 %>%
  filter(manu_site == "TX2" & JDAY_EXP > 1353)
```
#phe_hyp <- c("cgdd_12c_10d", "cgdd_12c_18d", "cgdd_12c_5d", 
#             "tave_5d", "tave_10d", "tave_18d")
phe_hyp <- c("FL50", "GR50", "dyln_fl50", "dyln_change_sec", "cgdd_12c_gr2fl", 
             "crain_gr2fl", "crain_1d", "crain_3d", "crain_5d")

```{r}
gr_ten_sites %>%
  saveRDS(file = here("data", "Weather_related_phe_GR50_3yr.rds"))
fl_ten_sites %>%
  saveRDS(file = here("data", "Weather_related_phe_FL50_3yr.rds"))
```

Make phenotypic correlation matrices for these new years, and new weather variables.

```{r}
metadata <- read_rds(here("data", "metadata.rds"))
subpop_v2 <- list(Gulf_and_Midwest = c("Gulf", "Midwest"), Midwest = "Midwest",
                  Gulf = "Gulf")
year_v = c(2019, 2020, 2021)

fl_ten_sites <- fl_ten_sites %>%
  left_join(select(metadata, PLANT_ID, SUBPOP))
gr_ten_sites <- gr_ten_sites %>%
  left_join(select(metadata, PLANT_ID, SUBPOP))

# change based on phenotype ---------------
#phe_hyp <- c("cgdd_12c_10d", "cgdd_12c_20d", "cgdd_12c_5d", 
#             "tave_5d", "tave_10d", "tave_20d")
phe_hyp <- c("FL50", "GR50", "dyln_fl50", "dyln_change_sec", 
             "dyln_change_sec_7d_prior", "dyln_change_sec_14d_prior", 
             "dyln_7d_prior", "dyln_14d_prior", "cgdd_12c_gr2fl", 
             "crain_gr2fl", "crain_1d", "crain_3d", "crain_5d")
i=1
k=1

U_hyp_2019 <- list()
for(i in seq_along(phe_hyp)){
  for(k in seq_along(subpop_v2)){

    cov_df <- fl_ten_sites %>%
      ungroup() %>%
      filter(SUBPOP %in% subpop_v2[[k]]) %>%
      filter(Year %in% 2019) %>%
      select(PLANT_ID, SUBPOP, manu_site, phe_hyp[i]) %>%
      group_by(PLANT_ID, SUBPOP, manu_site) %>%
      mutate(IND = row_number()) %>%
      pivot_wider(names_from = manu_site, values_from = phe_hyp[i]) %>%
      mutate(PLANT_ID = as.factor(PLANT_ID),
             IND = as.factor(IND)) %>%
      select(PLANT_ID, SUBPOP, IND, IL, MI, MO, NE, SD, TX1, TX2, TX3) #####

    cor_phe <- cor(cov_df[,-(1:3)], use = "pairwise")
    # Set diagonal of this matrix = coefficient of variation within each garden.
    diag(cor_phe) <- cov_df %>%
      ungroup() %>%
      summarise(IL = sqrt(var(IL, na.rm = TRUE))/mean(IL, na.rm = TRUE),  #####
                MI = sqrt(var(MI, na.rm = TRUE))/mean(MI, na.rm = TRUE),  #####
                MO = sqrt(var(MO, na.rm = TRUE))/mean(MO, na.rm = TRUE),
                NE = sqrt(var(NE, na.rm = TRUE))/mean(NE, na.rm = TRUE),
                SD = sqrt(var(SD, na.rm = TRUE))/mean(SD, na.rm = TRUE),
                TX1 = sqrt(var(TX1, na.rm = TRUE))/mean(TX1, na.rm = TRUE),
                TX2 = sqrt(var(TX2, na.rm = TRUE))/mean(TX2, na.rm = TRUE),
                TX3 = sqrt(var(TX3, na.rm = TRUE))/mean(TX3, na.rm = TRUE)) %>%
      as_vector()

    U_hyp_2019 <- list.prepend(U_hyp_2019, cor_phe)
    names(U_hyp_2019)[1] <- paste0(phe_hyp[i], "_", names(subpop_v2)[k])
  }
}

# for 2020
U_hyp_2020 <- list()
for(i in seq_along(phe_hyp)){
  for(k in seq_along(subpop_v2)){

    cov_df <- fl_ten_sites %>%
      ungroup() %>%
      filter(SUBPOP %in% subpop_v2[[k]]) %>%
      filter(Year %in% 2020) %>%
      select(PLANT_ID, SUBPOP, manu_site, phe_hyp[i]) %>%
      group_by(PLANT_ID, SUBPOP, manu_site) %>%
      mutate(IND = row_number()) %>%
      pivot_wider(names_from = manu_site, values_from = phe_hyp[i]) %>%
      mutate(PLANT_ID = as.factor(PLANT_ID),
             IND = as.factor(IND)) %>%
      select(PLANT_ID, SUBPOP, IND, IL, MI, MO, NE, SD, TX1, TX2, TX3, TX4) #####

    cor_phe <- cor(cov_df[,-(1:3)], use = "pairwise")
    # Set diagonal of this matrix = coefficient of variation within each garden.
    diag(cor_phe) <- cov_df %>%
      ungroup() %>%
      summarise(IL = sqrt(var(IL, na.rm = TRUE))/mean(IL, na.rm = TRUE),  #####
                MI = sqrt(var(MI, na.rm = TRUE))/mean(MI, na.rm = TRUE),  #####
                MO = sqrt(var(MO, na.rm = TRUE))/mean(MO, na.rm = TRUE),
                NE = sqrt(var(NE, na.rm = TRUE))/mean(NE, na.rm = TRUE),
                SD = sqrt(var(SD, na.rm = TRUE))/mean(SD, na.rm = TRUE),
                TX1 = sqrt(var(TX1, na.rm = TRUE))/mean(TX1, na.rm = TRUE),
                TX2 = sqrt(var(TX2, na.rm = TRUE))/mean(TX2, na.rm = TRUE),
                TX3 = sqrt(var(TX3, na.rm = TRUE))/mean(TX3, na.rm = TRUE),
                TX4 = sqrt(var(TX4, na.rm = TRUE))/mean(TX4, na.rm = TRUE)) %>%
      as_vector()

    U_hyp_2020 <- list.prepend(U_hyp_2020, cor_phe)
    names(U_hyp_2020)[1] <- paste0(phe_hyp[i], "_", names(subpop_v2)[k])
  }
}

U_hyp_2021 <- list()
for(i in seq_along(phe_hyp)){
  for(k in seq_along(subpop_v2)){

    cov_df <- fl_ten_sites %>%
      ungroup() %>%
      filter(SUBPOP %in% subpop_v2[[k]]) %>%
      filter(Year %in% 2021) %>%
      select(PLANT_ID, SUBPOP, manu_site, phe_hyp[i]) %>%
      group_by(PLANT_ID, SUBPOP, manu_site) %>%
      mutate(IND = row_number()) %>%
      pivot_wider(names_from = manu_site, values_from = phe_hyp[i]) %>%
      mutate(PLANT_ID = as.factor(PLANT_ID),
             IND = as.factor(IND)) %>%
      select(PLANT_ID, SUBPOP, IND, IL, MI, MO, TX2, TX3, TX4) #####

    cor_phe <- cor(cov_df[,-(1:3)], use = "pairwise")
    # Set diagonal of this matrix = coefficient of variation within each garden.
    diag(cor_phe) <- cov_df %>%
      ungroup() %>%
      summarise(IL = sqrt(var(IL, na.rm = TRUE))/mean(IL, na.rm = TRUE),  #####
                MI = sqrt(var(MI, na.rm = TRUE))/mean(MI, na.rm = TRUE),  #####
                MO = sqrt(var(MO, na.rm = TRUE))/mean(MO, na.rm = TRUE),
                TX2 = sqrt(var(TX2, na.rm = TRUE))/mean(TX2, na.rm = TRUE),
                TX3 = sqrt(var(TX3, na.rm = TRUE))/mean(TX3, na.rm = TRUE),
                TX4 = sqrt(var(TX4, na.rm = TRUE))/mean(TX4, na.rm = TRUE)) %>%
      as_vector()

    U_hyp_2021 <- list.prepend(U_hyp_2021, cor_phe)
    names(U_hyp_2021)[1] <- paste0(phe_hyp[i], "_", names(subpop_v2)[k])
  }
}

Visualize correlation matrices themselves
```{r}
library(switchgrassGWAS)

i=7
switchgrassGWAS::mash_plot_pairwise_sharing(corrmatrix = U_hyp_2019[[i]],
                                            reorder = FALSE)
switchgrassGWAS::mash_plot_pairwise_sharing(corrmatrix = U_hyp_2020[[i]],
                                            reorder = FALSE)
switchgrassGWAS::mash_plot_pairwise_sharing(corrmatrix = U_hyp_2021[[i]],
                                            reorder = FALSE)
```

How related are correlation matrices? Mantel tests & visualization of pairwise tests.
```{r}
for(i in seq_along(U_hyp_2019)) {
  U_hyp_2019[[i]] <- replace_na(U_hyp_2019[[i]], 0)
}

U_dist_2019 <- list()
for(i in seq_along(U_hyp_2019)) {
  U_dist_2019[[i]] <- as.dist(U_hyp_2019[[i]])

}
names(U_dist_2019) <- names(U_hyp_2019)

pval_mantel_2019 <- data.frame()
for(i in seq_along(U_dist_2019)) {
  for (j in seq_along(U_dist_2019)) {
      pval_mantel_2019[i, j] <- mantel.rtest(U_dist_2019[[i]], U_dist_2019[[j]], nrepet = 1000)[5] 
  }
}
colnames(pval_mantel_2019) <- names(U_dist_2019)
```

```{r}
plot_mantel_2019 <- pval_mantel_2019 %>%
  add_column(matrix_row = names(U_dist_2019), .before = TRUE) %>%
  pivot_longer(cols = 2:ncol(pval_mantel_2019)+1, names_to = "matrix_col", values_to = "pvalue") %>%
  #filter(matrix_row != matrix_col) %>%
  arrange(pvalue) %>%
  ggplot(aes(x = matrix_col, y = matrix_row)) + 
  geom_tile(aes(fill = -log10(pvalue) > 2.5, color = -log10(pvalue)), 
            size = 3.5, 
            width = 0.7, height = 0.7) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)) +
  scale_fill_viridis_d(end = 0.95, option = "B") +
  scale_color_viridis_c(end = 0.8, option = "D")

save_plot(plot = plot_mantel_2019, filename = here("analysis", "hypothesis_matrices", "Weather_Uhyp_Mantel_pvalue_plot_2019.png"), base_height = 13, base_width = 15)
```

```{r}
for(i in seq_along(U_hyp_2020)) {
  U_hyp_2020[[i]] <- replace_na(U_hyp_2020[[i]], 0)
}

U_dist_2020 <- list()
for(i in seq_along(U_hyp_2020)) {
  U_dist_2020[[i]] <- as.dist(U_hyp_2020[[i]])

}
names(U_dist_2020) <- names(U_hyp_2020)

pval_mantel_2020 <- data.frame()
for(i in seq_along(U_dist_2020)) {
  for (j in seq_along(U_dist_2020)) {
      pval_mantel_2020[i, j] <- mantel.rtest(U_dist_2020[[i]], U_dist_2020[[j]], nrepet = 1000)[5] 
  }
}
colnames(pval_mantel_2020) <- names(U_dist_2020)
```

```{r}
plot_mantel_2020 <- pval_mantel_2020 %>%
  add_column(matrix_row = names(U_dist_2020), .before = TRUE) %>%
  pivot_longer(cols = 2:ncol(pval_mantel_2020)+1, names_to = "matrix_col", values_to = "pvalue") %>%
  #filter(matrix_row != matrix_col) %>%
  arrange(pvalue) %>%
  ggplot(aes(x = matrix_col, y = matrix_row)) + 
  geom_tile(aes(fill = -log10(pvalue) > 2.5, color = -log10(pvalue)), 
            size = 3.5, 
            width = 0.7, height = 0.7) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)) +
  scale_fill_viridis_d(end = 0.95, option = "B") +
  scale_color_viridis_c(end = 0.8, option = "D")

save_plot(plot = plot_mantel_2020, filename = here("analysis", "hypothesis_matrices", "Weather_Uhyp_Mantel_pvalue_plot_2020.png"), base_height = 13, base_width = 15)
```

```{r}
for(i in seq_along(U_hyp_2021)) {
  U_hyp_2021[[i]] <- replace_na(U_hyp_2021[[i]], 0)
}

U_dist_2021 <- list()
for(i in seq_along(U_hyp_2021)) {
  U_dist_2021[[i]] <- as.dist(U_hyp_2021[[i]])

}
names(U_dist_2021) <- names(U_hyp_2021)

pval_mantel_2021 <- data.frame()
for(i in seq_along(U_dist_2021)) {
  for (j in seq_along(U_dist_2021)) {
      pval_mantel_2021[i, j] <- mantel.rtest(U_dist_2021[[i]], U_dist_2021[[j]], nrepet = 1000)[5] 
  }
}
colnames(pval_mantel_2021) <- names(U_dist_2021)
```

```{r}
plot_mantel_2021 <- pval_mantel_2021 %>%
  add_column(matrix_row = names(U_dist_2021), .before = TRUE) %>%
  pivot_longer(cols = 2:ncol(pval_mantel_2021)+1, names_to = "matrix_col", values_to = "pvalue") %>%
  #filter(matrix_row != matrix_col) %>%
  arrange(pvalue) %>%
  ggplot(aes(x = matrix_col, y = matrix_row)) + 
  geom_tile(aes(fill = -log10(pvalue) > 2.5, color = -log10(pvalue)), 
            size = 3.5, 
            width = 0.7, height = 0.7) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)) +
  scale_fill_viridis_d(end = 0.95, option = "B") +
  scale_color_viridis_c(end = 0.8, option = "D")

save_plot(plot = plot_mantel_2021, filename = here("analysis", "hypothesis_matrices", "Weather_Uhyp_Mantel_pvalue_plot_2021.png"), base_height = 13, base_width = 15)
```
